How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2024-09-11 2.17a-courses.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Choosing your courses

As a postgraduate student, your university should offer you a number of courses designed for postgraduate students. In many countries, these postgraduate courses will be compulsory and admission to the graduate school (Admission to Candidacy) will only be made after you have completed sufficient courses with grades to qualify you. For example, in the USA there are normally two years of these postgraduate courses followed by three years of working on your thesis project.

@@

1. There will be some courses that align directly to the contents of your thesis and you should do them
1. There should be a [writing course](https://johnmeasey.github.io/Thesis-Writing-Guide/) (and if not you can follow a link to my own) that you should attend as everyone needs help with writing
1. A time management course would also be very helpful, if your university has one
1. Look back at your short and medium term goals (see [here](#Plan)) and consider which courses could be relevant in your future
1. Do some courses just because you are interested. Given that you have already done a lot of courses at university, you are going to need some that hold genuine interest for you. Remember that diversity is an important aspect of science, so think as widely as possible.

@@

Some universities allow their students to attend postgraduate courses at other (usually nearby and affilitated) universities. If you know of a course that it being conducted at another university, but that you would like to attend, it might be worth approaching your postgraduate teaching co-ordinator to ask them if you can go.

There is also the opportunity to conduct a postgraduate course online. Again if you know of one, or put in the time to find an appropriate one, it may be worth approaching your postgraduate teaching co-ordinator to ask if you can do it.

The most important consideration is whether or not your institution is able to use marks generated at another institution in their own system to assess your postgraduate courses. Even if they don't allow you, it's worth talking to them about it as it may help those that come after you to find the course that they want. Or it might help stimulate to offer something similar at your institution in future.

@@

You need to make sure that you get the most out of these courses.

Looking back at being an undergraduate, you may have been focussed on grades or shortcutting work on courses, but for postgraduates this must have changed.

## Getting the most our of your courses

Remember that it is not direct applicability that is of the only importance.

@@

Courses are designed for postgraduates and usually taught by the most experienced professors. Use your time with them to [increase your network](https://howtopublishscience.org/advisorless.html#networks).

@@

I you haven't already worked this out for yourself, learning is a real gift and the postgraduate courses you are offered at universities are probably the ultimate in their level of sophistication and knowledge transfer. Take this opportunity seriously and you should find it extremely rewarding.

>>>>

# Choosing your postgraduate courses

As a postgraduate student, your university should offer you a number of courses designed specifically for postgraduate students. In many countries, these postgraduate courses will be compulsory and admission to the graduate school (Admission to Candidacy) will only be made after you have completed sufficient courses with grades to qualify you. These are often made up in terms of credits that you gain for each course attended and/or pass. There are normally one or two years of these postgraduate courses followed by three years of working on your thesis project.

Postgraduate courses are widely acknowledged as the reason why postgraduate students attain great advancement in their studies. As proof of this, know that countries that have historically never had such courses have now introduced them. Despite all PhD students already having a wide range of skills from their undergraduate and MSc studies, the skills attained from postgraduate courses are designed to augment their knowledge base to a higher level so that they attain systematic expertise within the discipline, and understand the dynamics of the frontier. Hence, these courses should not only provide the basic skills for you to understand the methods of scientific research, but they should enable you to fast-track to the vanguard of international academic frontiers.

@@

1. There will be some courses that align directly to the contents of your thesis and you should do them
1. There should be a [writing course](https://johnmeasey.github.io/Thesis-Writing-Guide/) (and if not you can follow a link to my own) that you should attend as everyone needs help with writing
1. A time management course would also be very helpful, if your university has one
1. Look back at your short and medium term goals (see [here](#Plan)) and consider which courses could be relevant in your future
1. Some courses may have certification and be essential during your studies but also useful in your following career
1. An important aim during this period is to build your network, so take note of who is teaching the courses and be cognisant how they might fit into your future network. 
1. The courses that you attend as a postgraduate may well be listed on your certification. Some courses may well endear you to certain employers, so consider your medium to long term goals (see [here](#Plan)) 
1. Do some courses just because you are interested. Given that you have already done a lot of courses at university, you are going to need some that hold genuine interest for you. Remember that diversity is an important aspect of science, so think as widely as possible

### Mandatory courses

Universities are usually very strict about their rules, but these should also available to you and so it is well worth asking for a copy. The rules for your institution may change from year to year and so it will be important to make sure that you have the relevant version, and if possible get these from the person responsible. These rules may also include other important information regarding the examination process (see [Part 1](#criteria)).

The compulsory courses mainly refer to the participation of graduate students in academic activities, teaching practice and social practice. As the name suggests, these are mandatory for all students and so there is no need to consider any choices here. It may be incumbent on you to ensure that you have fulfilled all necessary credits for these courses. For this reason, I suggest that you are familiar with the current rules of your faculty.

### Elective courses

Although you have your mandatory courses decided for you, you also will likely have the opportunity to do some elective courses. If there are elective courses that are a close fit to your thesis topic, then clearly you should elect to take these courses. The advice here comes more for other points that you may want to consider if you have credits to make up and cannot decide what to do.

1. **Consider your timetable:** You will have a timetable that you need to work to and make sure that you can physically get to a course as well as fit it in with your other mandatory and elective choices.
1. **Where does the course happen?:** the courses may take place in another campus or online. This may suit you better, or prohibit you from attending.
1. **How many credits does the course carry?:** You will need to make up a total amount of credits to gain admission to your postgraduate school. You might need to trade-off the time required for a course and the credits that it gives.
1. **Consider something from another faculty:** If allowed in your institution, consider taking a course given in another faculty. Some might be very challenging if you have no undergraduate or MSc experience, but others could be very rewarding. 
    1. Social Sciences - very useful courses might include GIS courses in geography, 
    1. Social Sciences - many biological systems have important human components that can really only be solved using questionnaires
    1. Social Sciences - philosophy has some very important lessons for biologists, especially in relation to logic and hypothesis building
    1. Art - There is a lot to be said for having a creative elective. A surprisingly large part of the process of doing science is actually creative. If you can conduct a course that will stimulate the creative part of your learning, so much the better. 
    1. Medicine - _Homo sapiens_ has become an excellent model organism, and might be much better studied in the same areas that you are working in. 
1. **Something enjoyable** - the chances are that some of your mandatory courses are going to be less enjoyable, so it's fair to give yourself something more enjoyable to look forward to during your week.

### Certified courses

You may find that attending and passing some postgraduate courses will provide you with a certificate. For example, you may need to attend a course on laboratory animal management and welfare. Certificates from such courses may well be advantageous to you in your career following your postgraduate studies, making such certificated courses particularly attractive.

@@

Some universities allow their students to attend postgraduate courses at other (usually nearby and affiliated) universities. If you know of a course that it being conducted at another university, but that you would like to attend, it might be worth approaching your postgraduate teaching co-ordinator to ask them if you can go.

There may also be the opportunity to conduct a postgraduate course online. Again if you know of one, or put in the time to find an appropriate one, it may be worth approaching your postgraduate teaching co-ordinator to ask if you can do it. 
The most important consideration is whether or not your institution is able to use marks generated at another institution in their own system to assess your postgraduate courses. Even if they don't allow you, it's worth talking to them about it as it may help those that come after you to find the course that they want. Or it might help stimulate to offer something similar at your institution in future.

@@

You need to make sure that you get the most out of these courses. Your attitude toward each course will be an important aspect of how well you perform and enjoy the course. Hence, if you feel that one of your mandatory courses is a waste of time and you wish you didn't have to do it, you will be unlikely to get much out of this course. On the other hand, if you have a positive attitude towards all of your courses (mandatory and elective) you will likely find them much more interesting and useful to your time as a postgraduate student.

Looking back at being an undergraduate, you may have been focussed on grades or short-cutting work on courses, but for postgraduates this must have changed. You should understand by now that learning is a two-way process and that engaging with your teachers is a vital part of improving your understanding of a topic. Your teachers are likely to have an incredible depth of information on the topic that they are teaching you, and probing this knowledge will help to to appreciate the subject and help you garner new interest. Therefore you should engage with your teachers on these courses. Remember that building your network is an important aim for you during this time.

@@

Courses are designed for postgraduates and usually taught by the most experienced professors. Use your time with them to [increase your network](https://howtopublishscience.org/advisorless.html#networks). You should be using the courses to network among as many fellow postgraduate students as possible. There are likely to be people using your approaches and techniques in various other disciplines within your university, and making these connections can be very important in the future.

@@

I you haven't already worked this out for yourself, learning is a real gift and the postgraduate courses you are offered at universities are probably the ultimate in their level of sophistication and knowledge transfer. Remember that your attitude is key in this process, so be sure that you are mentally prepared to having a positive attitude to all of your courses. Take this opportunity seriously and you should find it extremely rewarding.

### 2024-09-11 3.10-Figures.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021language]. The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector). The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.

>>>>

Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2024language]. The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector). The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.

### 2024-09-11 index.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 
title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
csl: peerj.csl
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the biological sciences. It will guide you from deciding  whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
github-repo: yihui/bookdown-crc
graphics: yes
---

@@

This book started as a series of blog posts written by me for my students ([john.measey.com/blog](john.measey.com/blog)). I wrote them because I felt that I was repeating the same advice year after year. Instead of rewriting comments on manuscripts time after time, I decided to consolidate them and put them all in one place. Later, I started looking for particular problems that students were having, and then writing blog posts to try to preempt those problems, and to provide solutions before tasks were completed. Finally, I have tried to fill in the missing gaps in the blog posts. But there are still plenty of gaps, and I regard this book as a work-in-progress. It is my intention that this book will always be added to, and improved by a community of practice, and I hope that this will include you.

@@

Embarking on a PhD is intimidating as, for most students, it will be their first experience working within the academic system. The voyage of discovery is often made very frustrating as much of what goes on in academia is assumed knowledge. Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated? What is needed is a guide that postgraduate students can refer to before, during and while making decisions about their time within academia. Note that this is not a rulebook. There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more. The intention then is to provide you with a starting point from which you can establish your confidence in the academic writing process, and build your own creativity.

@@

This book is intended to be a guide. Don't sit down and read it from cover to cover. If you do, then I think you have likely misunderstood what this guide is for. Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city, country or even like Douglas Adams' fabled electronic guide _The Hitchhiker's Guide to the Galaxy_ @adams2017ultimate (if you are not familiar [see here](https://en.wikipedia.org/wiki/The_Hitchhiker%27s_Guide_to_the_Galaxy)). However, instead of having hostels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.

@@

[**Part 1**](#part1)
-	Part one is written as something that you should read early on in the process, once you’ve committed to doing your PhD. This section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it. Perversely, it asks you what you want to do after you’ve finished, pointing out that you will have opportunities to open doors during the process. At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider. Working within academia is particularly stressful environment, and students should be aware of the state of their mental health, how to watch become aware if it is deteriorating, and what you can do to help yourself. I also encourage you to participate in the **scientific project**, and become aware of your role as a scientist within society.

@@

[**Part 4**](#part4)
-	The last and shortest section of this book talks about what you still need to do once your data chapters are written. For most PhD theses, you will need to write a short introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end. These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies. I have framed their content at the bare minimum required, as many students will find themselves under a great deal of pressure to submit their thesis to a prescribed deadline. I also provide some alternative approaches for those who have the luxury of more time to attack these sections in more detail. The very end of the book sets out some reasons for why you need to publish your thesis chapters.

Throughout this guide you will find links and references of things to read elsewhere. The reality is that there is a lot out there for you to read. I am not the only person to write a guide, and I can't pretend that all the answers are here in this book. You will find that in your particular situation you need to look around. What I aim to do in this book is provide you with enough background information to help in your onward search, and provide you with enough inside knowledge to do a PhD.

@@

So, this guide is my practical attempt to help you to get the most that you can out of your PhD time. After all, it should be an amazing journey that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations. When you have a PhD, you become a scientist. Too often, however, it's a nightmare journey of cul-de-sacs, and groping in the dark.

@@

@@

Before I go on, there are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who are exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially #academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.

>>>>

--- 
title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
csl: peerj.csl
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the biological sciences. It will guide you from deciding  whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
github-repo: https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences
graphics: yes
---

@@

This book started as a series of blog posts written by me for my students ([john.measey.com/blog](john.measey.com/blog)). I wrote them because I felt that I was repeating the same advice year after year. Instead of rewriting comments on manuscripts time after time, I decided to consolidate them and put them all in one place. Later, I started looking for particular problems that students were having, and then writing blog posts to try to pre-empt those problems, and to provide solutions before tasks were completed. Finally, I have tried to fill in the missing gaps in the blog posts. But there are still plenty of gaps, and I regard this book as a work-in-progress. It is my intention that this book will always be added to, and improved by a community of practice, and I hope that this will include you.

@@

Embarking on a PhD is intimidating as, for most students, it will be their first experience working within the academic system. The voyage of discovery is often made very frustrating as much of what goes on in academia is assumed knowledge. Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated? What is needed is a guide that postgraduate students can refer to before, during and while making decisions about their time within academia. Note that this is not a rulebook. Your university, faculty and department will all have rule books that you should obtain and become familiar with. This book does not replace these rule books. Instead, this guide tries to explain how to achieve your goals of getting a PhD. There are times when this guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more. The intention then is to provide you with a starting point from which you can establish your confidence in the academic writing process, and build your own creativity.

@@

This book is intended to be a guide. Originally, I called it the 'Fool's Guide to Writing a PhD in Biological Sciences', but that didn't fly. Don't sit down and read it from cover to cover. If you do, then I think you have likely misunderstood what this guide is for. Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city or country. However, instead of having hotels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.

@@

[**Part 1**](#part1)
-	Part one contains core skills that are relevant to all postgraduate students. These core skills are those that you will need to build upon from the very beginning of your studies. You are likely to already have some experience in some of them, but understanding how they fit together will provide a solid basis for your PhD process. Both critical reading and regular writing are core skills that you will need to develop from the outset of your studies. One of the most important aspects of doing a PhD is knowing what you want to use it for once you've completed it. Hence, this section of the book asks you to look not only at short term day-to-day planning, but also what you want to do after you’ve finished, pointing out that you will have opportunities to open doors during the duration of your PhD. At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider. Working within academia is particularly stressful, and students should be aware of the state of their mental health, how to measure it and determine whether it is deteriorating, and if so, what you can do to help yourself. I also encourage you to participate in the scientific project, and become aware of your role as a scientist within society.

@@

[**Part 4**](#part4)
-	The last and shortest section of this book talks about what you still need to do once your data chapters are written. For most PhD theses, you will need to write a short introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end. These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies. I have framed their content at the bare minimum required, as many students will find themselves under a great deal of pressure to submit their thesis to a prescribed deadline. I also provide some alternative approaches for those who have the luxury of more time to attack these sections in more detail. The very end of the book sets out some reasons why you need to publish your thesis chapters.

Throughout this guide you will find links and references of articles to read elsewhere. The reality is that there is a lot out there for you to read. I am not the only person to write a guide, and I can't pretend that all the answers are here in this book. You will find that in your particular situation you need to look around. What I aim to do in this book is provide you with enough background information to help in your onward search, and provide you with enough inside knowledge to do a PhD.

@@

So, this guide is my practical attempt to help you to get the most that you can out of your PhD time. You will doubtless hear many academics referring to your PhD studies as the best of times. Their retrospective view probably won't resonate with you as it's easy to forget all of the hard work involved. But you can only make the most out of your time if you know what you are trying to achieve, and that is the aim of this guide. After all, yoru PhD should be an amazing journey of discovery that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations. When you have a PhD, you become a scientist.

@@

## What's not in this book? {-}

Depending on just how early you are in your career, there is a lot missing from this book that has been provided in another book: [How to publish in Biological Sciences](https://howtopublishscience.org/index.html). That book concentrates on publishing manuscripts, while this book concentrates on getting students started writing data chapters. Hence, if you want extra information about publishing in the biological sciences, I would point you to the other book. If you are still writing and generating content then this is the right book for you.

If you want more help with publishing, and want help to demystify the publishing process, then [How to publish in Biological Sciences](https://howtopublishscience.org/index.html) has the content that you need:

```{r fig.align='centre', out.width = '40%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/'}
knitr::include_graphics('figures/cover2.png', dpi = NA)
```

There is common ground in both books, and I will point to important chapters relevant to publishing in the other book that are not reproduced here.

@@

Before I go on, there are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who are exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially #academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to Thalassa Matthews, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Ellie Brown, Sebastian Chekunov, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.
